A Transdiciplinary approach to developing a spectral numerical coastal-flow

### 2019-09-01 literature_review.Rmd
some additions to measuring waves section

Waves are complex and various approaches, techniques and devices have been developed in order to measure waves effectively. One of the ways that has been used extensivley used in the past but less so today, are visual estimates which can be used to characterise *significant swell height* (Hs) and *significant swell period* (Tp). Visual estimates use 15 - 20 of the most well defined, higher waves of a number of wave groups to characterise Hs and Tp. Although these parameters are useful, they do not accurately reflect the waves that are occuring in nature. Ocean waves are a combination of wind sea (short, irregular, locally generated waves) and swell (long, smooth waves, generated by distant storms) and so more parameters are needed to seperate Hs and Tp driven by wind sea and swell, i.e. seperate Hs and Tp parameters for wind sea and swell. However, even with seperate parameters for the different types of waves would still not be enough to effectivley characterise waves in the ocean. In order to characterise the detail and complexity of ocean waves a different approach must be used, know as the *spectral* approach. This approach is based on the idea that the sea surface can be characterised as the summation of a large number of harmonic wave components, known as the *random-phase/amplitude* model.

The wave spectrum is based on several spectra, namely the amplitude spectrum, variance spectrum, variance density spectrum (discontinous) and the variance density spectrum (continuous). A wave spectrum can be used to describe the sea surface as a stochastic process, however two conditions (assumptions) must first be met 1) the observation must be stationary and 2) the surface elevations must be Gaussian distributed. As mentioned previously, the random-phase/amplitude model approach characterises waves as harmonic and the time-record can be reproduced as the sum of a large number of harmonic waves using a *Fourier analysis* method. A Fourier analysis allows values of the amplitude and phase for each frequency to be characterised, which in turn provides the amplitude and phase spectrum for a single time-record. The amplitude spectrum is can be calculated by averaging many time-records which determines the *average amplitude spectrum*. This approach allows the entire wave record to be characterised but is not statistically relevant and therefore is not applicable to hydrodynamic modelling. In order for to attain a statistically relevant value, the variance of each wave component must be calculated. The variance is regarded as statistically relevant as the sum of the variances (i.e. random surface elevation) of the wave components is equal to the variance of the sum of the wave components. In addition, LWT dictates that the energy of waves is proportional to the variance and is therefore an indirect measure of wave energy. This indirect measurment can also be used to determine other important wave components such as wave-induced particle velocity and pressure variations. Important to note that the measurement of variance are discrete and is therefore limited in characterising waves. A continious scale is more applicable for hydrodynamic modelling purposes and is also closer to representing ocean waves and can be calculated by allowing the frequency interval () to approach zero.

@@

Wave exposure may be modelled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorporates integrative measures of hydrodynamic conditions at a particular site. Cartographical models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurements for larger areas, and has been suggested as a method for predicting macroalgal community structure (???). An example of such a model is the “BioEx model” which was developed by Baardseth and others (1970) to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiversity (???). Lindegarth and Gamfeldt (2005) critized this approach, arguing that the choice of wave exposure method can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices (Lindegarth and Gamfeldt 2005). Other authors, such as Hill et al. (2010), have argued that these simple measures can be improved upon by including bathymetry data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by Isæus (2004), and is known as the “simplified wave model” (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithms to mimic diffraction.

@@

This distribution follows a temperature gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, *E. maxima*, is distributed from Lüderitz to Cape Agulhas (Fig. 1) (Bolton and Levitt 1985, Probyn and McQuaid 1985, Bolton and Anderson 1987, Bolton et al. 2012). Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters (Bolton and Anderson 1987). There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role (Bustamante and Branch 1996). The estimated productivity of *E. maxima* within South Africa varies between 350 and 1500g Cm-2yr-1 (Mann 1982). Across the majority of the coastline, Laminaria pallida remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters (Field et al. 1980a, Bolton and Anderson 1987, Molloy and Bolton 1996). This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters (Field et al. 1980a, Molloy and Bolton 1996, Stegenga et al. 1997). Towards the north along the west coast, from around Hondeklipbaai, *L. pallida* replaces *E. maxima* as the dominant kelp species (Velimirov et al. 1977,Stegenga et al. 1997) and it also occupies increasingly shallow subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions (Molloy and Bolton 1996). This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as Laminaria schinzii Foslie (Molloy and Bolton 1996). Genetic work has subsequently shown that the two morphs are in fact the same species (Rothman et al. 2017). In southern African waters, the primary production of *Laminaria pallida* is between 120 and 1900g C m2yr1, similar to that of *E. maxima* (Mann 1982). Primary production is not the only pathway.

>>>>

Waves are complex and various approaches, techniques and devices have been developed in order to measure waves effectively. One of the ways that has been used extensively used in the past but less so today, are visual estimates which can be used to characterise *significant swell height* (Hs) and *significant swell period* (Tp). Visual estimates use 15 - 20 of the most well defined, higher waves of a number of wave groups to characterise Hs and Tp. Although these parameters are useful, they do not accurately reflect the waves that are occurring in nature. Ocean waves are a combination of wind sea (short, irregular, locally generated waves) and swell (long, smooth waves, generated by distant storms) and so more parameters are needed to separate Hs and Tp driven by wind sea and swell, i.e. separate Hs and Tp parameters for wind sea and swell. However, even with separate parameters for the different types of waves would still not be enough to effectively characterise waves in the ocean. In order to characterise the detail and complexity of ocean waves a different approach must be used, know as the *spectral* approach. This approach is based on the idea that the sea surface can be characterised as the summation of a large number of harmonic wave components, known as the *random-phase/amplitude* model. The random-phase amplitude model is a summation of wave components on a discrete scale, however a continous scale is more relevant to characterising and measuring of waves in the ocean. The discrete spectrums are converted to continoius spectrums through various methods. Therefore the random-phase amplitude model is used to obtain a continious variance density spectrum.

The wave spectrum is based on several spectra, namely the amplitude spectrum, variance spectrum, variance density spectrum (discontinuous) and the variance density spectrum (continuous). A wave spectrum can be used to describe the sea surface as a stochastic process, however two conditions (assumptions) must first be met 1) the observation must be stationary and 2) the surface elevations must be Gaussian distributed. As mentioned previously, the random-phase/amplitude model approach characterises waves as harmonic and the time-record can be reproduced by the sum of a large number of harmonic waves using a *Fourier analysis*. A Fourier analysis allows values of the amplitude and phase for each frequency to be characterised, which in turn provides the amplitude and phase spectrum for a single time-record. The amplitude spectrum is can be calculated by averaging many time-records which determines the *average amplitude spectrum*. This approach allows the entire wave record to be characterised but is not statistically relevant and therefore is not applicable to hydrodynamic modelling. In order for to attain a statistically relevant value, the variance of each wave component must be calculated. The variance is regarded as statistically relevant as the sum of the variances (i.e. random surface elevation) of the wave components is equal to the variance of the sum of the wave components. In addition, LWT dictates that the energy of waves is proportional to the variance and is therefore an indirect measure of wave energy. This indirect measurement can also be used to determine other important wave components such as wave-induced particle velocity and pressure variations. Important to note that the measurement of variance are discrete and is therefore limited in characterising waves. A continuous scale is more applicable for hydrodynamic modelling purposes and is also closer to representing ocean waves and can be calculated by allowing the frequency interval () to approach zero.

@@

Wave exposure may be modeled through various methods which range from simple cartographic to more advanced numerical wave models. Traditional ecological measures of wave exposure usually incorporates integrative measures of hydrodynamic conditions at a particular site. Cartographic models can be qualitative or quantitative and were designed for the need of wave exposure measures to explain ecological distributions. A simple set of calculations on coastline and wind data, and relatively small input data sets are required. These are regarded as "fetch-based models", which measure the length of open water associated with a site along a straight line. The output of such an approach is a simplified estimate of the potential wave energy for a specific set of sites. Advances in cartographical methods using fetch-based models has allowed for wave exposure measurements for larger areas, and has been suggested as a method for predicting macroalgal community structure (???). An example of such a model is the “BioEx model” which was developed by Baardseth and others (1970) to estimate wave exposure over large regions. BioEx requires frequency, strength and direction of winds, weighted by degree of exposure within various directions. BioEx is calculated as the sum of the index developed at different spatial scales (local, fjord and open). This method has been used in mapping of marine coastal biodiversity (???). Lindegarth and Gamfeldt (2005) criticised this approach, arguing that the choice of wave exposure method can influence ecological inference. The authors also highlighted the need for objective, reproducible and quantitative studies comparing exposure indices (Lindegarth and Gamfeldt 2005). Other authors, such as Hill et al. (2010), have argued that these simple measures can be improved upon by including bathymetry data which allows the incorporation of diffraction into the calculation. Diffraction is topographically induced variations in wave direction. A model incorporating this complexity was developed by Isæus (2004), and is known as the “simplified wave model” (SWM). The model uses measurements of wind strength, fetch and empirically derived algorithms to mimic diffraction.

@@

This distribution follows a temperature gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, *E. maxima*, is distributed from Lüderitz to Cape Agulhas (Fig. 1) (Bolton and Levitt 1985, Probyn and McQuaid 1985, Bolton and Anderson 1987, Bolton et al. 2012). Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters (Bolton and Anderson 1987). There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role (Bustamante and Branch 1996). The estimated productivity of *E. maxima* within South Africa varies between 350 and 1500g Cm-2yr-1 (Mann 1982). Across the majority of the coastline, Laminaria pallida remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters (Field et al. 1980a, Bolton and Anderson 1987, Molloy and Bolton 1996). This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters (Field et al. 1980a, Molloy and Bolton 1996, Stegenga et al. 1997). Towards the north along the west coast, from around Hondeklipbaai, *L. pallida* replaces *E. maxima* as the dominant kelp species (Velimirov et al. 1977,Stegenga et al. 1997) and it also occupies increasingly shallow subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions (Molloy and Bolton 1996). This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as *Laminaria schinzii* Foslie (Molloy and Bolton 1996). Genetic work has subsequently shown that the two morphs are in fact the same species (Rothman et al. 2017). In southern African waters, the primary production of *Laminaria pallida* is between 120 and 1900g C m2yr1, similar to that of *E. maxima* (Mann 1982). Primary production is not the only pathway.
